Association between SARS-CoV-2 and metagenomic content of samples from the Huanan Seafood Market

### 2023-07-02 paper/reviewer_response.tex

Related to the above point, it seems to me that the manuscript missed an opportunity to drive home the point that the data cannot be used to infer viral source.  The one way that one could get a smoking gun from this kind of data is if there was a sample that had a whole lot of viral material, and basically only one species associated with it (which didn’t happen).  In fact, the two samples with the highest SC2 levels didn’t have high levels of raccoon dog, human or any other susceptible species.  Using that as a foil one could point out that the positive with the most raccoon dog had only once SC2 copy, while the samples with the most SC2 copies ($>$1000 times more) had no prevalent candidate hosts associated with it at all.  Just a thought.

\response{
I agree that the main finding is that the metagenomic content of the environmental samples cannot be used to infer the viral source.
I have tried to emphasize this point as the major finding.
For instance, the abstract concludes with the statement that the samples are unable to ``reliably indicated whether any animals were infected by SARS-CoV-2.''
Likewise, the last two paragraphs of the discussion touch on this point (eg, ``When considered in the larger context, the inability of the environmental samples to inform on the origins of the virus is unsurprising.'').
However, I do not think it was \textit{a priori} certain that the samples would be uninformative on this question: as the reviewer notes, if there were a very strong association with a single species, that would be close to a smoking gun.

Overall, I agree with the reviewer's points above, and hope the current text, particularly the last two paragraphs of the discussion, adequately discuss these points.
}

### 2023-07-02 paper/reviewer_response.tex

\subsection*{Reviewer: 2}

Comments to the Author

In this paper, Bloom performs an independent analysis of metagenomic environmental samples from the Huanan Seafood market obtained through Chinese surveillance efforts. Because of the extensive press coverage and vigorous debates on social media regarding various claims and counter-claims about whether or not these data can be used to argue that potential animal intermediate hosts of SARS-CoV-2 (e.g. infected raccoon dogs) were present in the market, Bloom's analysis is valuable. Without this context, it would not be remarkable or publishable. However, it provides a very important null result, and also highlights a number of questionable scientific practices and overzealous interpretations by others. It should be therefore be published. I also commend Bloom on his unfailing commitment to reproducibility, with all code, and results made available (and usable)

\response{
I agree that this paper essentially just shows that these samples don't tell us much about the origin of the virus.
Normally, that would not be a finding of particular interest, but because there has been so much intense speculation about this question this paper is an example where an inconclusive or null result (simply showing that the data can't answer the question) is of interest.

I am glad the reviewer found the code and plots useful, and am hopeful that others have also used them to explore these data more.
}

@@

I have several specific comments/suggestions which I believe could strengthen the manuscript and improve its readability

1. I think it is critically important to specifically state that EVEN IF there was strong positive correlation between mammalian read counts and SC-2 read counts, it would still not be possible to use these data to argue that ANY particular animal was infected. Correlation $\ne$ causation, and there are too many confounders to note.

\response{
I am not certain agree with this point (see also the last point made by Reviewer 1 and my response).

I fully agree that these particular data aren't sufficient to tell either way whether any animal was infected.
But one can hypothetically imagine a scenario in which there was an extremely strong correlation between SARS-CoV-2 content and genetic material from a single species.
For instance, if all the samples with appreciable SARS-CoV-2 were dominated by raccoon dog material and no other samples had any SARS-CoV-2, that would in my view be strong evidence for infected raccoon dogs.
Likewise if all the samples with appreciable SARS-CoV-2 were dominated by human genetic material it would strongly suggest humans were shedding the virus.
Of course, in actuality it turns out that there isn't any trend like that, and to the extent there is any correlation it is with species that don't make sense as having been infected.
So I fully agree that it turns out that these samples are not sufficient to show that any particular animal was infected, and there are plenty of reasons we can see in hindsight why this is the case (samples collected late, low viral content, lots of mixing of material from different species, samples collected at different times, etc).

However, I do think that hypothetically if there had been very strong patterns in the data it could have provided strong evidence for infected animals of a particular species, so I don't really feel comfortable stating that the samples could not have been informative even in principle.
}

### 2023-07-02 paper/reviewer_response.tex

However, I do think that hypothetically if there had been very strong patterns in the data it could have provided strong evidence for infected animals of a particular species, so I don't really feel comfortable stating that the samples could not have been informative even in principle.
}

2. There is very robust literature on estimating species abundance from RNASeq-type metagenomic data. Specialized tools (e.g. Kraken 2) exist for efficient and accurate classification of such reads. It would be worthwhile to explain briefly why it is necessary to develop a "de novo" pipeline. Many commonly recognized issues (e.g. short reads which are not diagnostic at species level because they are found in multiple genomes) are well understood and quantified. C-C (presumably) used contigs to address this lack of ``read-level'' precision (they are very SHORT in the data discussed), but that's a hack.

\response{
This is an excellent question.
My primary reason for using the \textit{de novo} mitochondrial pipeline rather than something like Kraken2 was to enhance the comparability of my results to those of Crits-Christoph et al, since I thought a major question was likely to be if any discrepancies between their findings and mine were due to methodological differences.
Crits-Christoph et al did not make the relevant code for the metagenomic analyses publicly available alongside their report, so I wrote a \textit{de novo} pipeline that mirrored what they described in the methods of their report.

There are also several technical reasons why standard use of Kraken2 with off-the-shelf reference genome sets would not work:
\begin{itemize}
\item A standard Kraken2 pre-built genomic database would not be suitable for this study, as some of the species of particular interest (eg, bamboo rat) do not even have sequenced genomes on RefSeq.
\item Use of mitochondrial rather than nuclear genomes is therefore probably preferable because a much wider range of species, including essentially all the potentially susceptible non-human species at the market, have fully sequenced mitochondrial genomes.
\item Although this is not clearly explained in the methods of Crits-Christoph et al, at the current time at least one of the species that is a focus of their study (Amur hedgehog) does not even have a mitochondrial genome on RefSeq, and that sequence had to be obtained elsewhere manually from Genbank (see methods of my pre-print). Therefore, it required some steps beyond just downloading RefSeq mitochondrial genomes to build a reference set comparable to that used by Crits-Christoph et al.
\item Given these complexities, I wanted to be sure I explicitly understood how all steps were being handled in reference building and handling of multi-mapping reads. This made me prefer a pipeline where I had written each step explicitly rather than a package like Kracken2 which streamlines several steps together.
\end{itemize}

Nonetheless, I think a similar analysis could have been performed using Kraken2, and I mostly took the approach here to ensure as much comparability as possible to the report by Crits-Christoph et al.

I have added text to the portion of the manuscript that explains the methods used for this part of the analysis explaining the points above.
}

### 2023-07-02 paper/paper.tex
Reviewer 2 point 2: methods rationale

To create an mitochondrial genome alignment reference set, I followed a procedure partially analogous to that described by \citet{crits2023genetic}.
All mitochondrial genomes in the RefSeq database were downloaded, and then filtered to retain only genomes from the phylum Chordata (in contrast, \citet{crits2023genetic} describe retaining all metazoa mitochondrial genomes).
The reason I limited to chordate rather than metazoa mitochondrial genomes is that all chordates have very similar length mitochondrial genomes, but at the level of metazoa there is wide variation in mitochondrial genome length.
The Amur hedgehog mitochondrial genome was not present in this set, so it was separately downloaded from Genbank accession KX964606.
All the mitochondrial genomes were then filtered to remove highly similar ones.
To do this, \texttt{Mash}~\citep{ondov2016mash} was used to compute the mash distances between all pairs of mitochondrial genomes.
To ensure the most relevant mitochondrial genomes were retained, I first specified for manual retention the genomes for all the species listed in the third supplementary file of \citet{crits2023genetic}, as well as genomes for some additional relevant species (these species are listed under the \textit{mitochondrial\_genomes\_to\_keep} key in \url{https://github.com/jbloom/Huanan_market_samples/blob/main/config.yaml}).
After retaining all of these manually specified mitochondrial genomes, the pipeline then greedily iterated through all other genomes choosing for retention each genome that was not within a mash distance of 0.07 of another already retained genome.
The full set of retained mitochondrial genomes is listed in Table~\ref{tab:mito_retained}.

>>>>

To create an mitochondrial genome alignment reference set, I followed a procedure partially analogous to that described by \citet{crits2023genetic}.
Note that a similar analysis could also be implemented using more standard metagenomic analysis tools such as Kraken2~\citep{wood2019improved}, as was in fact done in the Chinese CDC study~\citep{liu2023surveillance}.
The reasons for instead implementing the pipeline used here were as follows:
\begin{itemize}
\item To ensure comparability to \citet{crits2023genetic} by using an approach that parallels the one they described.
\item To use a reference set that includes all of the potentially susceptible species sold at the market: many of these species lack full nuclear genomes on RefSeq, and one (Amur hedgehog) doesn't even have a mitochondrial genome on RefSeq. Therefore, the manual process used to build the mitochondrial reference set described below was necessary to ensure all these species were in the reference.
\end{itemize}

For the current study, all mitochondrial genomes in the RefSeq database were downloaded, and then filtered to retain only genomes from the phylum Chordata (in contrast, \citet{crits2023genetic} describe retaining all metazoa mitochondrial genomes).
The reason I limited to chordate rather than metazoa mitochondrial genomes is that all chordates have very similar length mitochondrial genomes, but at the level of metazoa there is wide variation in mitochondrial genome length.
The Amur hedgehog mitochondrial genome was not present in this set, so it was separately downloaded from Genbank accession KX964606.
All the mitochondrial genomes were then filtered to remove highly similar ones.
To do this, \texttt{Mash}~\citep{ondov2016mash} was used to compute the mash distances between all pairs of mitochondrial genomes.
To ensure the most relevant mitochondrial genomes were retained, I first specified for manual retention the genomes for all the species listed in the third supplementary file of \citet{crits2023genetic}, as well as genomes for some additional relevant species (these species are listed under the \textit{mitochondrial\_genomes\_to\_keep} key in \url{https://github.com/jbloom/Huanan_market_samples/blob/main/config.yaml}).
After retaining all of these manually specified mitochondrial genomes, the pipeline then greedily iterated through all other genomes choosing for retention each genome that was not within a mash distance of 0.07 of another already retained genome.
The full set of retained mitochondrial genomes is listed in Table~\ref{tab:mito_retained}.
